\pkg{hesim}: Health Economic Simulation Modeling and Decision Analysis

### 2021-02-13 manuscript.Rnw
Some minor tweaks

  Jeroen Jansen\\
  University of California, San Francisco\\
  San Francisco\\
  E-mail: \email{jeroen.jansen@ucsf.edu}\\
}

@@

\section{Introduction} \label{sec:intro}
Health technology assessment (HTA) is a systematic approach for comparing competing health technologies to inform the efficient use of health care resources. Publicly funded health systems such as those in Australia, Canada, and the United Kingdom, among others, use HTA to help maximize health gains for the population given a fixed budget. HTA is also commonly used by private payers to guide coverage decisions and the adoption of new technologies \citep{trosman2011health}. Such assessments usually rely heavily on cost-effectiveness analysis (CEA) so that decisions can be made on the basis of a formal evaluation of costs and effects \citep{dakin2015influence}. CEA is made feasible by development of health economic models that simulate costs and effects over relevant time horizons using the totality of available evidence.

@@

Expected values are computed by integrating the ``weighted'' probability of being in each state, where weights are a function of the discount factor and state values. That is, for a time horizon $T$, discounted costs and QALYs in health state h are computed as,

@@

\subsection{Individual continuous time state transition models} \label{sec:iCTSTMs}
iCTSTMs simulate individual trajectories between health states using random number generation. Trajectories are simulated for multiple patients and costs and QALYs are computed by averaging across the simulated patients. A reasonably large number of patients must be simulated to ensure that expected values are stable \citep{ohagan2007monte}.

@@

Survival functions are estimated by fitting parametric or flexible parametric survival models as described in Section~\ref{sec:iCTSTMs}. The $n$th fitted survival model with density $f_n(t)$ has cumulative density function $F_n(t)$, survivor function $1 - F_n(T)$, cumulative hazard $H_n(t) = -\log S_n(t)$, and hazard $h_n(t) = f_n(t)/S_n(t)$. State probabilities for each health state, $x_{hc}$, can be computed for an arbitrarily fine grid of time periods $c$. Costs and QALYs are then computed in the same manner as in the cohort model described in Section~\ref{sec:cDTSTMs}.

%% -- Framework ----------------------------------------------------------------
\section{Framework} \label{sec:framework}
Economic models consist of a disease model, a utility model, and a set of cost models for each cost category. Model development proceeds in 4 steps. The first step is to setup the model by defining the model structure, target population, and treatment strategies of interest. The relevant information is then stored in a \code{hesim_data} object.

Development and analysis of the model is performed in the next three steps as shown in Figure~\ref{fig:econ-eval-process-hesim}. First, the disease progression, costs, and utility submodels are parameterized using ``estimation'' datasets. Next, the submodels are combined to construct an economic model. The economic model is then used to simulate, disease progression, QALYs, and costs as a function of ``input data''. The input data always contains variables describing the target population and treatment strategies of interest, but can also contain variables related to time to facilitate use of time-varying covariates or parameters. Finally, the simulated QALYs and costs are used to perform decision analysis within a cost-effectiveness framework. While other approaches such as multi-criteria decision analysis (MCDA) could, in principle, be used as well, only CEA is currently supported. All models are simulated using PSA so that decision uncertainty can be represented.

@@

The three economic models described in Section~\ref{sec:model-taxonomy} are implemented as the \pkg{R6} classes \code{CohortDtstm}, \code{IndivCtstm}, and \code{Psm}. Each class contains methods for simulating disease progression, QALYs, and costs. These methods are made possible by separate \pkg{R6} classes for the disease, utility, and cost submodels. The remainder of this section describes the framework for parameterization and simulation. The cost-effectiveness framework is described in Section~\ref{sec:cea}.

@@

\subsection{Simulation} \label{sec:simulation}
The utility and cost models are always \pkg{R6} objects of class \code{StateVals}, whereas the disease models vary by economic model (Table~\ref{tbl:submodels}). The disease model is used to simulate survival curves in a PSM and health state transitions in a cDTSTM and iCTSTM.

\begin{table} [h]
\caption{Submodels comprising an economic model}\label{tbl:submodels}
\footnotesize
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}l l l l}
\hline
Economic model & Disease model & Utility model & Cost model(s)\\
\hline
\code{CohortDtstm} & \code{CohortDtstmTrans} & \code{StateVals} & \code{StateVals}\\
\code{Psm} & \code{PsmCurves} & \code{StateVals} & \code{StateVals}\\
\code{IndivCtstm} & \code{IndivCtstmTrans} & \code{StateVals} & \code{StateVals}\\
\hline
\end{tabular*}
\end{table}

@@

Each economic model contains methods for simulating disease progression, QALYs, and costs (Table~\ref{tbl:simulate-outcomes}). The utility and cost models always simulate QALYs and costs from the simulated progression of disease with the methods \code{$sim_qalys()} and \code{$sim_costs()}, respectively. Methods for simulating disease progression differ slightly since the simulation approaches differ as described in Section~\ref{sec:model-taxonomy}. In an N-state PSM $n-1$ survival curves are generated and in a iCTSTM a trajectory through the STM is simulated for multiple patients via individual simulation. All models can simulate state probabilities over time.

\begin{table} [h]
\caption{Methods for simulating outcomes from economic models}\label{tbl:simulate-outcomes}
\footnotesize
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}l l l l}
\hline
Economic model & Disease progression  & QALYs & Costs \\
\hline
\code{CohortDtstm} & \code{sim_stateprobs()} & \code{sim_qalys()} & \code{sim_costs()}\\
\code{Psm} & \code{sim_survival()}, \code{sim_stateprobs()}  & \code{sim_qalys()} & \code{sim_costs()}\\
\code{IndivCtstm} & \code{sim_disease()}, \code{sim_stateprobs()}  & \code{sim_qalys()} & \code{sim_costs()}\\
\hline
\end{tabular*}
\end{table}

@@

The complete economic model is an constructed by combining the transition, utility, and cost submodels.

>>>>

  Jeroen P Jansen\\
  University of California, San Francisco\\
  San Francisco\\
  E-mail: \email{jeroen.jansen@ucsf.edu}\\
}

@@

\section{Introduction} \label{sec:intro}
Health technology assessment (HTA) is a systematic approach for comparing competing health technologies to inform the efficient use of health care resources. Publicly funded health systems such as those in Australia, Canada, and the United Kingdom, among others, use HTA to help maximize health gains for the population given a fixed budget. HTA is also used by private payers to guide coverage decisions and the adoption of new technologies \citep{trosman2011health}. Such assessments usually rely heavily on cost-effectiveness analysis (CEA) so that decisions can be made on the basis of a formal evaluation of costs and effects \citep{dakin2015influence}. CEA is made feasible by development of health economic models that simulate costs and effects over relevant time horizons for the compared health technologies using the totality of relevant evidence for the decision problem at hand.

@@

Expected values are computed by integrating the probability of being in each state and ``weighted'' state values, where weights are a function of the discount factor. That is, for a time horizon $T$, discounted costs and QALYs in health state h are computed as,

@@

\subsection{Individual continuous time state transition models} \label{sec:iCTSTMs}
iCTSTMs simulate individual trajectories between health states using random number generation. Trajectories are simulated for multiple patients and expected costs and QALYs are computed by averaging across the simulated patients. A reasonably large number of patients must be simulated to ensure that expected values are stable \citep{ohagan2007monte}.

@@

Survival functions are estimated by fitting (flexible) parametric survival models as described in Section~\ref{sec:iCTSTMs}. The $n$th fitted survival model with density $f_n(t)$ has cumulative density function $F_n(t)$, survivor function $1 - F_n(T)$, cumulative hazard $H_n(t) = -\log S_n(t)$, and hazard $h_n(t) = f_n(t)/S_n(t)$. State probabilities for each health state, $x_{hc}$, can be computed for an arbitrarily fine grid of time periods $c$. Costs and QALYs are then computed in the same manner as in the cohort model described in Section~\ref{sec:cDTSTMs}.

%% -- Framework ----------------------------------------------------------------
\section{Framework} \label{sec:framework}
Economic models with \pkg{hesim} consist of a disease model, a utility model, and a set of cost models for each cost category. Model development proceeds in 4 steps. The first step is to setup the model by defining the model structure, target population, and treatment strategies of interest. The relevant information is then stored in a \code{hesim_data} object.

Development and analysis of the model is performed in the next three steps as shown in Figure~\ref{fig:econ-eval-process-hesim}. First, the disease progression, costs, and utility submodels are parameterized using ``estimation'' datasets. Next, the submodels are combined to construct an economic model. The economic model is then used to simulate, disease progression, QALYs, and costs as a function of ``input data''. The input data always contains variables describing the target population and treatment strategies of interest, but can also contain variables related to time to facilitate use of time-varying covariates or parameters. Finally, the simulated QALYs and costs are used to perform decision analysis within a cost-effectiveness framework. While other approaches such as multi-criteria decision analysis (MCDA) could, in principle, be used as well, only CEA is currently supported. All models are simulated using PSA so that parameter uncertainty can be represented.

@@

The three types of economic models described in Section~\ref{sec:model-taxonomy} are implemented as the \pkg{R6} classes \code{CohortDtstm}, \code{IndivCtstm}, and \code{Psm}. Each class contains methods for simulating disease progression, QALYs, and costs. These methods are made possible by separate \pkg{R6} classes for the disease, utility, and cost submodels. The remainder of this section describes the framework for parameterization and simulation. The cost-effectiveness framework is described in Section~\ref{sec:cea}.

@@

\subsection{Simulation} \label{sec:simulation}
The utility and cost models are always \pkg{R6} objects of class \code{StateVals}, whereas the disease models vary by economic model (Table~\ref{tbl:submodels}). The disease model is used to simulate health state transitions in a cDTSTM and iCTSTM, and survival curves in a PSM.

\begin{table} [h]
\caption{Submodels comprising an economic model}\label{tbl:submodels}
\footnotesize
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}l l l l}
\hline
Economic model & Disease model & Utility model & Cost model(s)\\
\hline
\code{CohortDtstm} & \code{CohortDtstmTrans} & \code{StateVals} & \code{StateVals}\\
\code{IndivCtstm} & \code{IndivCtstmTrans} & \code{StateVals} & \code{StateVals}\\
\code{Psm} & \code{PsmCurves} & \code{StateVals} & \code{StateVals}\\
\hline
\end{tabular*}
\end{table}

@@

Each economic model contains methods for simulating disease progression, QALYs, and costs (Table~\ref{tbl:simulate-outcomes}). The utility and cost models always simulate QALYs and costs from the simulated progression of disease with the methods \code{$sim_qalys()} and \code{$sim_costs()}, respectively. Methods for simulating disease progression differ slightly since the simulation approaches differ as described in Section~\ref{sec:model-taxonomy}. In a iCTSTM a trajectory through the STM is simulated for multiple patients via individual simulation and in an N-state PSM $n-1$ survival curves are generated and . All models can simulate state probabilities over time.

\begin{table} [h]
\caption{Methods for simulating outcomes from economic models}\label{tbl:simulate-outcomes}
\footnotesize
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}l l l l}
\hline
Economic model & Disease progression  & QALYs & Costs \\
\hline
\code{CohortDtstm} & \code{sim_stateprobs()} & \code{sim_qalys()} & \code{sim_costs()}\\
\code{IndivCtstm} & \code{sim_disease()}, \code{sim_stateprobs()}  & \code{sim_qalys()} & \code{sim_costs()}\\
\code{Psm} & \code{sim_survival()}, \code{sim_stateprobs()}  & \code{sim_qalys()} & \code{sim_costs()}\\
\hline
\end{tabular*}
\end{table}

@@

The complete economic model is a constructed by combining the transition, utility, and cost submodels.
